Supplementary Information

### 2021-03-10 doc/manuscript/ForestGEO_treerings.Rmd
work on JTM comments

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].

@@

*(keep mostly the same, but shift emphasis more towards ecology):*
Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is very challenging and uncertain.
Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_age-related_2011]. 
However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.

@@

We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`. 
Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016]. 
We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).

@@

Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by the approach.

@@

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the most important climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual cores, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.

@@

There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017] 
With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig. `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.

>>>>

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling at sites with characteristics that result in trees being more climate-limited, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].

@@

*(keep mostly the same, but shift emphasis more towards ecology):*
Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is very challenging and uncertain.
Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors. 
Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc. -- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_age-related_2011]. 
However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.

@@

We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for four species, as detailed in Appendix `r traditional_comparison`. 
Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016]. 
We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).

@@

Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig. `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of the drivers of tree growth in an era of global change.
Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by the approach.

@@

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the most important climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged. 
Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.

@@

There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017]. 
With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig. `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
